### 2013-09-25 manuscripts/dpp-msbayes/dpp-msbayes.tex
Adding more text to model description in dpp ms.

\subsubsection*{Priors on divergence models}
\begin{linenomath}
The prior term for the vector of divergence times for each of
the \npairs{} pairs of populations can be expanded as
\begin{equation}
    p(\divTimeMapVector) = p(\divTimeIndexVector)p(\divTimeVector \given \divTimeIndexVector),
    \label{eq:divModelPrior}
\end{equation}
where \divTimeVector is an ordered set of unique divergence time parameters
$\{\divTime{1}, \ldots, \divTime{\divTimeNum}\}$ whose length
\divTimeNum can range from 1 to \npairs{},
and \divTimeIndexVector is vector of indices of divergence time parameters 
$(\divTimeIndex{1}, \ldots, \divTimeIndex{\npairs{}})$
that map the divergence time in \divTimeVector to each of the \npairs{}
pairs of populations.
Thus, \divTimeMapVector is the result of applying the mapping function
\begin{equation}
    f(\divTimeVector, \divTimeIndexVector, i) = \divTime{\divTimeIndex{i}}
    \label{eq:divTimeMapFunction}
\end{equation}
to each population pair $i$, such that
$\divTimeMapVector = (\divTimeMap{1} = f(\divTimeVector, \divTimeIndexVector,
1), \ldots, \divTimeMap{\npairs{}} = f(\divTimeVector, \divTimeIndexVector,
\npairs{}))$.
Each \divTime{} within \divTimeVector is \iid as
$\divTime{} \sim Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
\end{linenomath}

Biologically speaking, \divTimeVector contains the times of divergence events,
the length of which \divTimeNum is the number of divergence events shared across
the \npairs{} pairs of populations.
For example, if \divTimeVector contains a single divergence time parameter
\divTime{1}, all \npairs{} pairs of populations are constrained to diverge at
this time (i.e., \divTimeIndexVector would contain the index 1 repeated
\npairs{} times, and \divTimeMapVector would contain the value \divTime{1}
repeated \npairs{} times), whereas if it contains \npairs{} divergence time
parameters, the model is fully generalized to allow all of the pairs to diverge
at unique times.

\begin{linenomath}
The total sample space of ordered realizations of \divTimeMapVector is all the
possible partitions of \npairs{} elements into 1 to \npairs{} categories.
The total number of possible partitions is a sum of the Stirling numbers of
the second kind over all possible number of categories \divTimeNum
\begin{equation}
    B_{\npairs{}}=\sum_{\divTimeNum=1}^{\npairs{}} \left[
    \frac{1}{\divTimeNum!} \sum_{j=0}^{\divTimeNum-1} (-1)^{j}
    \binom{\divTimeNum}{j} (\divTimeNum-j)^{\npairs{}} \right],
    \label{eq:bell}
\end{equation}
which is the Bell Number \citep{Bell1934}.
The original \msb model only samples over the unordered realizations of
\divTimeIndexVector, such that the sample space is reduced to all the possible
integer partitions of \npairs{} \citep{Oaks2012,OeisPartitionNumber,
    OeisPartitionTriangle,Malenfant2011}.
This has the advantage of reducing the sample space of the number of
divergence models, but has two disadvantages.
First, additional information about the data is lost.
By throwing away the identity of the \npairs{} pairs of populations, all
information regarding the amount of data (e.g., the number gene copies
collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), the
information content in the data (i.e., the amount of sequence variation), as
well as the taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) for each pair is lost.
The original descriptions of the \msb model claim that the independence
between the sample sizes \popSampleSize{1}{}, \popSampleSize{2}{} and
the summary statistic $\pi_b$ calculated from the alignment for each
population pair allows exchangeability.
This is incorrect.
The other summary statistics, and the alignments they are based on are
not independent of the sample sizes.
The alignments and summary statitics are not identically distributed because of
taxon- and locus-specific parameters \hkyModel{}{},
\mutationRateScalarConstant{}{}, and \ploidyScalar{}{}
Second, the identity of the pairs of populations is lost.
\end{linenomath}

We implement two prior probability distributions over the joint space of
\divTimeVector, \divTimeIndexVector.
One simply places a uniform prior over all possible integer paritions of
\npairs{}.

>>>>

\subsubsection*{Priors on divergence models}
\begin{linenomath}
The prior term for the vector of divergence times for each of
the \npairs{} pairs of populations can be expanded as
\begin{equation}
    p(\divTimeMapVector) = p(\divTimeIndexVector)p(\divTimeVector \given \divTimeIndexVector),
    \label{eq:divModelPrior}
\end{equation}
where \divTimeVector is an ordered set of unique divergence time parameters
$\{\divTime{1}, \ldots, \divTime{\divTimeNum}\}$ whose length
\divTimeNum can range from 1 to \npairs{},
and \divTimeIndexVector is vector of indices of divergence time parameters 
$(\divTimeIndex{1}, \ldots, \divTimeIndex{\npairs{}})$
that map the divergence time in \divTimeVector to each of the \npairs{}
pairs of populations.
Thus, \divTimeMapVector is the result of applying the mapping function
\begin{equation}
    f(\divTimeVector, \divTimeIndexVector, i) = \divTime{\divTimeIndex{i}}
    \label{eq:divTimeMapFunction}
\end{equation}
to each population pair $i$, such that
$\divTimeMapVector = (\divTimeMap{1} = f(\divTimeVector, \divTimeIndexVector,
1), \ldots, \divTimeMap{\npairs{}} = f(\divTimeVector, \divTimeIndexVector,
\npairs{}))$.
\end{linenomath}

Biologically speaking, \divTimeVector contains the times of divergence events,
the length of which \divTimeNum is the number of divergence events shared across
the \npairs{} pairs of populations.
For example, if \divTimeVector contains a single divergence time parameter
\divTime{1}, all \npairs{} pairs of populations are constrained to diverge at
this time (i.e., \divTimeIndexVector would contain the index 1 repeated
\npairs{} times, and \divTimeMapVector would contain the value \divTime{1}
repeated \npairs{} times), whereas if it contains \npairs{} divergence time
parameters, the model is fully generalized to allow all of the pairs to diverge
at unique times.
Regardless of the number of divergence time parameters (\divTimeNum), each
\divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.

\begin{linenomath}
The total sample space of ordered realizations of \divTimeMapVector is all the
possible partitions of \npairs{} elements into 1 to \npairs{} categories.
The total number of possible partitions is a sum of the Stirling numbers of
the second kind over all possible number of categories \divTimeNum
\begin{equation}
    B_{\npairs{}}=\sum_{\divTimeNum=1}^{\npairs{}} \left[
    \frac{1}{\divTimeNum!} \sum_{j=0}^{\divTimeNum-1} (-1)^{j}
    \binom{\divTimeNum}{j} (\divTimeNum-j)^{\npairs{}} \right],
    \label{eq:bell}
\end{equation}
which is the Bell Number \citep{Bell1934}.
The original \msb model only samples over the unordered realizations of
\divTimeIndexVector, such that the sample space is reduced to all the possible
integer partitions of \npairs{} \citep{Oaks2012,Huang2011,OeisPartitionNumber,
    OeisPartitionTriangle,Malenfant2011}.
We denote the set of all possible integer partitions of \npairs{} as
\integerPartitionSet{\npairs{}} and the length of that set as
\integerPartitionNum{\npairs{}}.
The advantages, disadvantages, and justification of ignoring the order
of \divTimeIndexVector is discussed in detail below.
\end{linenomath}

\begin{linenomath}
We implement two prior probability distributions over the space
of all possible discrete divergence models (\divTimeIndexVector).
The first simply gives all possible unordered partitions of \npairs{} elements
equal probability
\begin{equation}
    p(\divTimeIndexVector) = frac{1}{\integerPartitionNum{\npairs{}}},
    \label{eq:divModelPriorUniform}
\end{equation}
i.e., a discrete uniform prior over all the integer partitions of \npairs{}
(discrete divergence models).
We denote this prior as
$\divTimeIndexVector \sim DU(\integerPartitionSet{\npairs{}})$.
\end{linenomath}

The second prior we implement is based on the Dirichlet-process, which is a
stochastic process that groups elements into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The process is controlled by the concentration parameter \concentrationParam,
which determines how clustered the process will be.
We use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} elements, which we denote as $\divTimeIndexVector \sim
DP(\concentrationParam)$.
We use a hierarchical Bayesian approach and place a second-order prior
probability distribution (i.e., hyperprior) on \concentrationParam.  More
specifically, we use a gamma-distributed prior $\concentrationParam =
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the user.

\begin{linenomath}
This provides a great deal of flexibility for specifying the prior uncertainty
regarding divergence models.
The \concentrationParam determines the prior probability that any two
pairs of populations $i$ and $j$ will be assigned to the same divergence time
parameter
\begin{equation}
    p(\divTimeIndex{i} = \divTimeIndex{j}) = frac{1}{1 + \concentrationParam},
    \label{eq:dppPriorSameClass}
\end{equation}
and also the prior probability of the number of divergence time parameters
\begin{equation}
    p(\divTimeNum \given \concentrationParam, \npairs{}) = 
    \frac{\stirlingFirst{\npairs{}}{\divTimeNum} \concentrationParam^{\divTimeNum}}
    {\prod_{i=1}^{\npairs{}}(\concentrationParam + i - 1)},
    \label{eq:dppPriorNumClasses}
\end{equation}
where \stirlingFirst{\cdot,\cdot} are the unsigned Stirling numbers of the
first kind.
Equations \ref{eq:dppPriorSameClass} and \ref{eq:dppPriorNumClasses} show that
smaller values of \concentrationParam will favor fewer divergence time
parameters, and thus more clustered models of divergence, whereas larger values
will favor more divergence time parameters, and thus less clustered models of
divergence.
\end{linenomath}

\subsection*{Differences between our model and the original \msb model}
\subsubsection*{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution placed on divergence models.
As discussed in \citet{Oaks2012}, in \msb the only prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (\divTimeNum here; \numt{} in \citet{Huang2011}) imposes
an odd U-shaped prior over discrete divergence models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides the user with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

\subsubsection*{The priors on nuisance parameters}
Following the recommendations of \citet{Oaks2012}, we have replaced the use of
inappropriate, continuous uniform distributions for priors on many of the
model's parameters (\divTime{}, \ancestralTheta{}, \descendantTheta{1}{},
\descendantTheta{2}{}, \bottleScalar{1}{}, \bottleScalar{2}{},
\recombinationRate, \migrationRate{}) with more flexible parametric
distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
Not only do these priors have nice conjugacy properties, but they also provide
the user with much greater flexibility in expressing the prior uncertainty
regarding the parameters of the model.

Another distinction between our model and that of \msb, is the prior on the
sizes of the descendant populations of each pair.
As described by \citet{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Dirichlet(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under our model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.

\subsubsection*{Flexibility in parameterizing the model}
We also provide the user with the ability to control the richness of
the model.
For the \myTheta{} parameters, the model can be fully generalized to
allow each population pair to have three parameters:
\ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}.
Furthermore, any model of \myTheta{} parameters nested within this
general model can also be specified, including the most restricted model
where the ancestral and descendant populations of each pair share
a single \myTheta{} parameter.
We also allow the user to eliminate the parameters associated with
the post-divergence bottlenecks in the descendant populations of
each pair
(\bottleTime{}, \bottleScalar{1}{}, and \bottleScalar{2}{}),
which constrains the descendant populations to be of
constant size from present back to the divergence event.
This allows the user to specify a model that has as many as
seven parameters per population pair
(\ancestralTheta{}, \descendantTheta{1}{}, \descendantTheta{2}{},
\bottleTime{}, \bottleScalar{1}{}, \bottleScalar{2}{}, and
\migrationRate{})
or as few as
1 parameters per pair
(\myTheta{}),
in addition to the $\popSampleSize{i}{} - 1$ coalescence times.

\subsection*{The ABC implementation of the model}

This has the advantage of reducing the sample space of the number of
divergence models, but has two disadvantages.
First, additional information about the data is lost.
By throwing away the identity of the \npairs{} pairs of populations, all
information regarding the amount of data (e.g., the number gene copies
collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), the
information content in the data (i.e., the amount of sequence variation), as
well as the taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) for each pair is lost.
The original descriptions of the \msb model claim that the independence
between the sample sizes \popSampleSize{1}{}, \popSampleSize{2}{} and
the summary statistic $\pi_b$ calculated from the alignment for each
population pair allows exchangeability.
This is incorrect.
The other summary statistics, and the alignments they are based on are
not independent of the sample sizes.
The alignments and summary statitics are not identically distributed because of
taxon- and locus-specific parameters \hkyModel{}{},
\mutationRateScalarConstant{}{}, and \ploidyScalar{}{}
Second, the identity of the pairs of populations is lost.
